---
title: "Detecting Flu Epidemics via Search Engine Quieries"
output: github_document
date: "2022-12-17"
---

# About the Study and the Data Set

Flu epidemics are a major concern for public health because they can lead to respiratory illnesses, hospitalizations, and deaths. According to a report published in 2012, influenza was the eighth leading cause of death in the United States in 2011. Each year, between 250,000 and 500,000 people around the world die from influenza-related diseases. The U.S. Centers for Disease Control and Prevention (CDC) and the European Influenza Surveillance Scheme (EISS) track influenza activity using data from virologic tests and clinical visits for Influenza-like Illness (ILI). However, national and regional data on influenza activity is usually published with a lag of 1-2 weeks. The Google Flu Trends project was created to see if it would be possible to get faster reporting on influenza activity by using data from online searches related to the flu, which is available almost immediately.

### EDA

The goal is to use Google search logs to estimate influenza-like illness (ILI) activity. The Centers for Disease Control and Prevention (CDC) publishes data on its website on a weekly basis showing the percentage of patient visits to healthcare providers for ILI purposes at the regional and state levels. Google Trends allows anyone to access weekly counts for every query searched by users around the world. For each location, the counts are normalized by dividing the count for each query in a particular week by the total number of online search queries submitted in that location during the week, and then adjusting the values to be between 0 and 1. The FluTrain.csv file contains data from January 1, 2004 to December 31, 2011 and includes columns for the date range, the percentage of ILI-related physician visits for that week, and the fraction of queries related to ILI for that week. To begin analyzing this data, it should be loaded into a data frame called FluTrain.

```{r}
#Loading the dataset
FluTrain = read.csv("FluTrain.csv")
#Figuring at what week is ILI max
subset(FluTrain, ILI == max(ILI))

```

First, we can see that the maximum ILI is 7.618 and happened on the week 18/10/2009

```{r}
#Now Let's see the skewness of ILI data
hist(FluTrain$ILI)

```

The ILI data is skewed to the right, meaning there exist more lower ILI values than higher ones in our data set.

```{r}
plot(FluTrain$Queries, FluTrain$ILI)
plot(FluTrain$Queries, log(FluTrain$ILI))
```

As we can see from the two scatter plots, there's seem to be a positive linear relation between the quires and logarithmic scale of the ILI. The reason we would check the logarithmic is because our ILI data is naturally skewed.

### Linear Regression Model

```{r}
lmodel = lm(log(ILI) ~ Queries, data = FluTrain)
summary(lmodel)
```

As we can see, $R^2$ is equal to 0.709, which is relatively good and suggests that our initial assumption that quires are correlated with the logarithmic scale of ILI is correct.\
Let's apply the model to the training test.

```{r}
FluTest = read.csv("FluTest.csv")
predTest = exp(predict(lmodel, newdata = FluTest))
```

Here, we applied the exponential to the prediction because we trained our model on the logarithmic. and we want our predictions to be in the form of ILI not Log(ILI).

So, Let's find the RMSE of the model,

```{r}
SSE = sum((predTest - FluTest$ILI)^2)
RMSE = sqrt(SSE/nrow(FluTest))
RMSE
```

So the RMSE of the model is 0.749

### Time Series

The dataset for this analysis contains weekly measurements of the dependent and independent variables. This type of dataset is called a "time series" because the observations are taken at consecutive intervals of time. In this case, the observations are taken on a weekly basis. In order to improve the statistical model, it may be helpful to use the value of the dependent variable from earlier weeks to predict the current value of the dependent variable. For this analysis, this means that the ILI variable for the current week will be predicted using values of the ILI variable from previous weeks. To do this, the observations need to be "lagged," or delayed, by a certain amount of time. Because the ILI variable is reported with a 1- or 2-week lag, the decision maker will only have data available from 2 or more weeks ago. To account for this lag, a variable called ILILag2 will be created that contains the ILI value from 2 weeks before the current observation.

```{r}
library(zoo)
ILIlag2 = lag(zoo(FluTrain$ILI), -2, na.pad = TRUE)
#na.pad = TRUE, adds "missing values" or NA's to the start of the data set
FluTrain$ILIlag2 = coredata(ILIlag2)
```

Now lets's check the missing values in the new variable.

```{r}
summary(FluTrain$ILIlag2)
```

The two missing values are the ones at the start of the records where we have no data about the ILI values.\
Now let's see the relation between ILI and ILIlag2

```{r}
plot(FluTrain$ILI, FluTrain$ILIlag2)
plot(log(FluTrain$ILI), log(FluTrain$ILIlag2))
```

As with the previous model, we are going to use the log of the new variable.

```{r}
lmodel_updated = lm(log(ILI) ~ Queries + log(ILIlag2), data = FluTrain)
summary(lmodel_updated)
```

As we can see, all the variables are of high significance and the $R^2$ value has increased to 0.9063.

### Testing the Updated Model

```{r}
ILIlag2 = lag(zoo(FluTest$ILI), -2, na.pad = TRUE)
FluTest$ILIlag2 = coredata(ILIlag2)
summary(FluTest$ILIlag2)
```

For this analysis, the training set contains all of the observations from 2004 to 2011, and the testing set contains all of the observations from 2012. The training and testing sets are split sequentially, with no time gap between them. This means that the first observation in the testing set (FluTest) was recorded one week after the last observation in the training set (FluTrain). Based on this information, it is possible to determine how to fill in the missing values for the ILILag2 variable in FluTest. Specifically, the ILILag2 value for the first observation in FluTest can be calculated using the ILI value from two weeks before the last observation in FluTrain. Similarly, the ILILag2 value for the second observation in FluTest can be calculated using the ILI value from two weeks before the second-to-last observation in FluTrain, and so on.

```{r}
nrow(FluTrain)
FluTest$ILIlag2[1] = FluTrain$ILI[416]
FluTest$ILIlag2[2] = FluTrain$ILI[417]
```

```{r}
summary(FluTest$ILIlag2)
```

No missing values :D.

```{r}
predTest_updated = exp(predict(lmodel_updated, newdata = FluTest))
```

Let's calculate the RMSE of the model.

```{r}
SSE = sum((predTest_updated - FluTest$ILI)^2)
RMSE = sqrt(SSE/nrow(FluTest))
RMSE
```

As we can see the $RMSE$ has dropped significantly to 0.294, which could indicate a huge improvement in the model.
